Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-06 ms.tex
remove huge space at begining of paragraphs
first version of modifier invasion proof

\documentclass[9pt, a4paper, twocolumn]{extarticle}
\usepackage[margin=1in]{geometry} % See geometry.pdf to learn the layout options.
\geometry{a4paper}
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

@@

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem*{theorem}{Theorem}
\newtheorem*{corollary}{Corollary}
\newtheorem*{lemma}{Lemma}
\newtheorem*{remark}{Remark}
\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
\newtheorem*{caswell}{Lemma: Caswell's Formula}
\newtheorem*{frv}{Fisher's reproductive values}
\renewcommand\qedsymbol{} % no square at end of proof

@@

% Title page
\title{Necessity is the mother of genetic invention}

@@

% Abstract
\begin{abstract}
Generation of variation is considered detrimental in well-adapted populations evolving under constant selection.
Therefore, in a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find a general and simple result:
if individuals with below-average fitness increase the rate at which variation is generated, then the population mean fitness increases.
This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
and even to the famous proverb: \emph{"Necessity is the mother of genetic invention"}.
\end{abstract}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism commonly called \emph{stress-induced mutagenesis}, in which the mutation rate increases in individuals with poor fitness.
\citet{Ram2012} demonstrated that modifiers that increase the mutation rate of individuals with below-average fitness actually \textbf{increase} the population mean fitness, rather than decrease it.
Their analysis assumed an infinite population size and constant-viability selection, such that the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes.
\citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to those of~\citet{Kimura1966a} and~\citet{Haigh1978}.

Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of individual, and specifically on the its fitness.
Our result shows that low fitness individuals should produce more variation than high fitness individuals.

@@

\paragraph*{General model.}
Consider a large haploid population with an arbitrary (possibly infinite) set of types $A_1, A_2, \ldots$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{\omega} f' = (I-C+MC)D f,
\end{equation}
where $f_k$ is the frequency of type $A_k$ ($f=(f_1, f_2, \ldots)$ is a stochastic vector: $\sum_k{f_k} = 1$);
$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_j {M_{i,j}} = 1$ for any column \emph{j}); 
$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type); 
$I$ is the identity matrix;
and $\bar{\omega}$ is a normalizing factor such that $\sum_k{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_i$, the distribution of the destination types $A_j$ is given by $M_{j,i}$ (note the index order).
We focus on a modifier that has no direct effect on fitness and whose sole function is to determine the rate of transition out of the different types ($C_k$).
Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$. The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.

@@

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_k{f_k}=1$.
Then after selection and mutation, $f'_k$ in the next generation is given by
\begin{equation}
f_k' = (1 - \mu_k) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{\mu_j \frac{w_j}{\bar{w}} x_j},
\label{eq:mutation_model_1}
\end{equation}
for all $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k=1}^{n}{w_k f_k}$ is the population mean fitness.

@@

Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype.
A modifier locus then determines the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles.
In this case, we assume a low mutation rate $m_k$ such that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a new mutant allele and moving the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, moving the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

@@

To rephrase this model in the terms of the general model, we define

\begin{equation}
C_k = m_k, \quad
D_k = w_k,
%M_{i,j} = \begin{cases}
%\beta, & i = j-1 \\
%1-\beta-\delta, & i = j \\
%\delta, & i = j+1 \\
%1-\delta, & i=j=0 \\
%0, & \text{otherwise}
%\end{cases}
\end{equation}
and
\begin{equation}
M_{i,j} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots \\
\delta & 1-\delta-\beta & \beta &  \ddots \\
0 & \delta & 1-\delta-\beta & \ddots \\
\vdots & \ddots & \ddots & \ddots 
\end{bmatrix}
\end{equation}
so that Eq.~\ref{eq:mutation_model_2} is equivalent to Eq.~\ref{eq:model}.

@@

Here we assume a large population of haploids that occupies two demes and a modifier locus that determine the migration rates between the demes, $m_k$. 
Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $u_k$ in deme 2 for $k=1,2,\ldots,n$.
Let the frequencies of $A_k$ in deme 1 be $x_k$ and in deme 2 $y_k$ for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$. and $\sum_{k=1}^{n}{y_k} = 1$.

Then following selection and migration, after one generation the new frequencies $x_k'$ and $y_k'$ are given by
\begin{align}
x_k' = (1-m_k) \frac{w_k}{\bar{w}} x_k + m_k \frac{u_k}{\bar{u}} y_k \\
y_k' = m_k \frac{w_k}{\bar{w}} x_k + (1 - m_k) \frac{u_k}{\bar{u}} y_k
\end{align}
for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{u} = \sum_{k \ge 0}{u_k y_k}$ is the population mean fitness in deme 1 and 2, respectively.
This is an extension of a model presented by~\citet{Altenberg2017}, allowing for the migration rate of an individual, $m_k$, to depend on the identity of the allele $A_k$.

To rephrase this model in terms of the general model, we define
\begin{equation}
\begin{aligned}
f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
C = diag[m_1, \ldots, m_n, m_1, \ldots, m_n], \\
D = diag[w_1, \ldots, w_n, u_1, \ldots, u_n], \\
M = \begin{bmatrix}
0 & I_n \\
I_n & 0
\end{bmatrix}
\end{aligned}
\end{equation}
where $I_n$ is the n-by-n identity matrix.

@@

% Mean fitness principle
With the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{\omega}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{\omega}^* v = (I-C+MC)D v
\end{equation}

The following theorem presents a \emph{mean fitness principle} for the
sensitivity of the stable population mean fitness $\bar{\omega}^*$ to changes
in $C_k$ the probability for transition out of $A_k$.

\begin{mfp}
Let $\bar{\omega}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
\begin{equation}\label{eq:theorem}
\frac{\partial \bar{\omega}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:sign_theorem}
sign\frac{\partial \bar{\omega}^*}{\partial C_k} = 
sign(\bar{\omega}^* - D_k), 
\end{equation}
and therefore increased transition from type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
\end{mfp}

\begin{proof}
By the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007},
$\bar{\omega}^*$ is the leading real positive eigenvalue of $(I-C+MC)D$,
and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{v_k} = 1$ and $\sum_k{u_k v_k} = 1$.

Using \emph{Caswell's formula} (Eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
\begin{equation}
\frac{\partial \bar{\omega}^*}{\partial C_k} = 
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v.
\end{equation}

@@

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
u\tr \bar{\omega}^* = u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\bar{\omega}^*$ and the $k$
element of $u$:
\begin{equation}
u_k \bar{\omega}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
\end{equation}
Multiplying both sides by $v_k$ (which is positive if $D_k>0$) and rearranging, we get 
\begin{equation}
\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k) = v_k D_k(u\tr [M]_k - u_k).
\end{equation}
Plugging this in Eq.~\ref{eq:theorem_intermediate} yields:
\begin{equation}
\frac{\partial \bar{\omega}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k).
\end{equation}
Finally, note that $u_k, v_k, C_k \ge 0$ and therefore
\begin{equation}
sign \frac{\partial \bar{\omega}^*}{\partial C_k} = 
sign (\bar{\omega}^* - D_k),
\end{equation}
which completes the proof.
\end{proof}

Note that if the modifier has the same effect on all types (i.e.,
$C_k = \alpha$ for all $k$) then Eq.~\ref{eq:model_equilibrium}
becomes $\bar{\omega}^* v = ((1-\alpha)I + \alpha M)Dv$,
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above proof, we get
$$
\frac{\partial \bar{\omega}^*}{\partial \alpha} = 
\bar{\omega}^* - 1 \le 0,
$$
because $\bar{\omega}^* = \sum_i{v_i D_i}$ (see~\nameref{sec:AppA}) and $D_i \le 1$ for all $i$.

@@

\begin{equation}
\begin{aligned}
\frac{\partial \bar{\omega}^*}{\partial C_k} = 
v_k D_k (u\tr[M]_k - u_k) = \\
D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{aligned}
\end{equation}
Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, which is the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$).

@@

\begin{rvp}
In the notation of the theorem,
\begin{equation}
sign \frac{\partial \bar{\omega}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
\end{equation}
and therefore increased transition out of type $k$ will increase the stable population mean fitness if and only if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{rvp}

@@

We turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, and a modifier that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
\end{equation}
where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types $K$ is
\begin{equation}
\begin{aligned}
\frac{\partial \bar{\omega}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in \mathbb{K{}}}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in \mathbb{K}}{u_k v_k (\bar{\omega}^* - D_k)},
\end{aligned}
\end{equation}
which can be simplified to
\begin{multline}\label{eq:sign_theorem_threshold}
sign \frac{\partial \bar{\omega}^*}{\partial \tau} = 
sign \sum_{k \in \mathbb{K}}{u_k v_k (\bar{\omega}^* - D_k)} = \\ 
sign \Big(\bar{\omega}^* - \frac{\sum_{k \in \mathbb{K}}{u_k v_k D_k}}{\sum_{k \in \mathbb{K}}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{eq:sign_theorem}.
Therefore, increased transition from types $K$ will increase the stable population mean fitness if and only if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.

>>>>

\documentclass[9pt, a4paper, twocolumn]{extarticle}
\usepackage[margin=1in]{geometry} % See geometry.pdf to learn the layout options.
\geometry{a4paper}
%\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

@@

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem*{theorem}{Theorem}
\newtheorem*{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem*{remark}{Remark}
\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
\newtheorem*{frv}{Fisher's reproductive values}
\renewcommand\qedsymbol{} % no square at end of proof

@@

% Title page
\title{Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention}

@@

% Abstract
\begin{abstract}
Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find a general and simple result:
if individuals with below-average fitness increase the rate at which variation is generated, then the population mean fitness increases.
This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
and even to the proverb: \emph{"Necessity is the mother of genetic invention"}.
\end{abstract}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves \emph{stress-induced mutagenesis}, a mechanism in which the mutation rate increases in individuals with poor fitness.
Empirical evidence suggests \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
\citet{Ram2012} demonstrated that modifiers that increase the mutation rate of individuals with below-average fitness actually \textbf{increase} the population mean fitness, rather than decrease it.
Their analysis assumed an infinite population size and constant-viability selection, where the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes.
A similar result has been demonstrated for fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.

\citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.

Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of individual, and specifically on the its fitness.
Our result shows that if below-average individuals produce more variation than above-average individuals, then the population mean fitness increases.

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary (possibly infinite) set of types $A_1, A_2, \ldots$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f_k$ is the frequency of type $A_k$ ($f=(f_1, f_2, \ldots)$ is a stochastic vector: $\sum_k{f_k} = 1$);
$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_j {M_{i,j}} = 1$ for all \emph{j}); 
$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type); 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_k{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).
We focus on a modifier that has no direct effect on fitness and whose sole function is to determine the rate of transition out of the different types ($C_k$).
Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$. The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.

@@

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_k{f_k}=1$.
Then after selection and mutation, $f'_k$ in the next generation is given by
\begin{equation}
f_k' = (1 - \mu_k) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{\mu_j \frac{w_j}{\bar{w}} f_j},
\label{eq:mutation_model_1}
\end{equation}
for all $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k=1}^{n}{w_k f_k}$ is the population mean fitness.

@@

Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype.
A modifier locus then determines the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles.
In this case, we assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

@@

In terms of the general model, we can write
\begin{equation}
C_k = m_k, \quad
D_k = w_k,
%M_{i,j} = \begin{cases}
%\beta, & i = j-1 \\
%1-\beta-\delta, & i = j \\
%\delta, & i = j+1 \\
%1-\delta, & i=j=0 \\
%0, & \text{otherwise}
%\end{cases}
\end{equation}
and
\begin{equation}
M_{i,j} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots \\
\delta & 1-\delta-\beta & \beta &  \ddots \\
0 & \delta & 1-\delta-\beta & \ddots \\
\vdots & \ddots & \ddots & \ddots 
\end{bmatrix},
\end{equation}
in which case Eq.~\ref{eq:mutation_model_2} is a specific case of Eq.~\ref{eq:model}.

@@

Here we assume a large population of haploids that occupies two demes and a modifier locus that determine the migration rates between the demes, $m_k$. 
Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $q_k$ in deme 2 for $k=1,2,\ldots,n$.
Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$, and $\sum_{k=1}^{n}{y_k} = 1$.

Then following selection and migration, after one generation the new frequencies $x_k'$ and $y_k'$ are given by
\begin{align}
x_k' = (1-m_k) \frac{w_k}{\bar{w}} x_k + m_k \frac{q_k}{\bar{q}} y_k \\
y_k' = m_k \frac{w_k}{\bar{w}} x_k + (1 - m_k) \frac{q_k}{\bar{q}} y_k
\end{align}
for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{q} = \sum_{k \ge 0}{q_k y_k}$ are the population mean fitnesses in deme 1 and 2, respectively.
This is an extension of a model presented by~\citet{Altenberg2017}, allowing for the migration rate of an individual, $m_k$, to depend on the identity of the allele $A_k$.

To rephrase this model in terms of the general model, we define
\begin{equation}
\begin{aligned}
f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
C = diag[m_1, \ldots, m_n, m_1, \ldots, m_n], \\
D = diag[w_1, \ldots, w_n, q_1, \ldots, q_n], \\
M = \begin{bmatrix}
0 & I_n \\
I_n & 0
\end{bmatrix}
\end{aligned}
\end{equation}
where $I_n$ is the n-by-n identity matrix.

% Modifier model
\paragraph*{Modifier model.}
We now consider the case of two modifier alleles, $m$ and $M$, in complete linkage to the types $A_k$.
These modifiers potentially induce different transition probabilities out of types $A_k$, with transition matrices $C=diag[C_1, \ldots, C_n]$ and $\tilde{C}=diag[\tilde{C}_1, \ldots, \tilde{C}_n]$.
The frequencies of type $A_k$ linked to modifier $m$ or $M$ is $f_k$ and $g_k$, respectively, such that $\sum_k{f_k+g_k}=1$;
$\bar{w}$ now ensures that $\sum_{k}{f_k'+g_k'}=1$;
the rest of the model parameters are the same as in Eq.~\ref{eq:model};
and the frequencies in the next generation, $f'$ and $g'$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Note that Eq.~\ref{eq:model} is a special case for $g_k=0$ for all $k$.
Here, $\bar{w}=\sum_{k}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).

@@

% Mean fitness principle
Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v
\end{equation}

The following theorem presents a \emph{mean fitness principle} for the
sensitivity of the stable population mean fitness $\bar{w}^*$ to changes
in $C_k$ the probability of transition from $A_k$.

\begin{mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
\begin{equation}\label{eq:theorem}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:sign_theorem}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

Therefore increased transition from type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
\end{mfp}

\begin{proof}
By the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007},
$\bar{w}^*$ is the leading real positive eigenvalue of $(I-C+MC)D$,
and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{v_k} = 1$ and $\sum_k{u_k v_k} = 1$.

Using \emph{Caswell's formula} (Eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v.
\end{equation}

@@

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
u\tr \bar{w}^* = u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\bar{w}^*$ and the $k$
element of $u$:
\begin{equation}
u_k \bar{w}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
\end{equation}
Multiplying both sides by $v_k$ (which is positive if $D_k>0$) and rearranging, we get 
\begin{equation}
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k) = v_k D_k(u\tr [M]_k - u_k),
\end{equation}
which when substituted into Eq.~\ref{eq:theorem_intermediate} yields:
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k).
\end{equation}
Finally, note that $u_k, v_k, C_k \ge 0$ and therefore
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (\bar{w}^* - D_k),
\end{equation}
which completes the proof.
\end{proof}

The above theorem provides a condition for the effect of a modifier in control of $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
Specifically, if $A_k$ individuals have below-average fitness, then a modifier that increases $C_k$ will increase the population mean fitness.

@@

\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial C_k} = 
v_k D_k (u\tr[M]_k - u_k) = \\
D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{aligned}
\end{equation}
Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$).

@@

\begin{rvp}
In the notation of the theorem,
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
\end{equation}
and therefore increased transition out of type $k$ will increase the stable population mean fitness if and only if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{rvp}

@@

We turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, and a modifier that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
\end{equation}
where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)},
\end{aligned}
\end{equation}
which can be simplified to
\begin{multline}\label{eq:sign_theorem_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{eq:sign_theorem}.
Therefore, increased transition from types $K$ will increase the stable population mean fitness if and only if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.

If such a modifier has the same effect on all individuals (i.e., $C_k = \alpha$ for all $k$) then Eq.~\ref{eq:model_equilibrium} $\bar{w}^* v = ((1-\alpha)I + \alpha M)Dv$,
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above proof, we find
$$
\frac{\partial \bar{w}^*}{\partial \alpha} = 
\bar{w}^* - 1 \le 0,
$$
because $\bar{w}^* = \sum_i{v_i D_i}$ (see~\nameref{sec:AppA}), and $D_i \le 1$ for all $i$.

% Evolutionary stability

We determined a condition for a modifier that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
But could such a modifier increase in frequency when initially rare in the population?
To answer this question we will analyze the stability of modifier $m$ with transition rates $C_k$ to the invasion of modifier $M$ with rates $\tilde{C}_k$, as described in Eq.~\ref{eq:modifier_model}.

The equilibrium of Eq.~\ref{eq:modifier_model} when $M$ is missing from the population ($g_k=0$ for all $k$) is (modified from \ref{eq:model_equilibrium})
\begin{equation}\label{eq:modifier_model_equilibrium}
\bar{w} \cdot \begin{pmatrix}v \\ 0\end{pmatrix} = 
\Big(\vec{I} - \vec{C} + \vec{M} \vec{C}\Big)\vec{D} \begin{pmatrix}v \\ 0\end{pmatrix},
\end{equation}
where $\vec{I}$ is a $2n$-by-$2n$ identity matrix, and
\begin{equation}
%\begin{aligned}
\vec{C} =  \begin{pmatrix}C & 0 \\ 0 & \tilde{C} \end{pmatrix}, \quad
\vec{M} =  \begin{pmatrix}M & 0 \\ 0 & M \end{pmatrix}, \quad
\vec{D} =  \begin{pmatrix}D & 0 \\ 0 & D \end{pmatrix}.
%\end{aligned}
\end{equation}

The stability of modifier $m$ to invasion by modifier $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $L_{ex}$ of modifier $M$, which is in turn determined by the Jacobian of the system in Eq.~\ref{eq:modifier_model_equilibrium}
\begin{equation}
J = \begin{pmatrix}L_{in} & 0 \\ 0 & L_{ex} \end{pmatrix}.
\end{equation}
Note the zero block matrices due to complete linkage between the modifier and the types $A_k$.

$L_{ex}$ can be written 
\begin{equation}\label{eq:Lex}
L_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
\end{equation}
where $\bar{w}^*$ is the stable population mean fitness without modifier $M$ (see~\nameref{sec:AppA}).
